---
title: "Lab 1 - Data Structures"
author: "Ali Zaidi"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
random_norms <- rnorm(100)
some_letters <- letters[1:10]
library(learnr)
```

# R Primitives

## Quick Tour of Things You Need to Know
### Data Structures

> "Bad programmers worry about the code. 
> Good programmers worry about data structures and their relationships."
> - Linus Torvalds

* R's data structures can be described by their dimensionality, and their type.


|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

### Data Types

* Atomic vectors come in one of four types
* `logical` (boolean). Values: `TRUE` | `FALSE`
* `integer`
* `double` (often called numeric)
* `character`
* Rare types:
* `complex` 
* `raw`

## Manipulating Data Structures

### Subsetting Operators

* To create a vector, use `c`: `c(1, 4, 1, 3)`
* To create a list, use `list`: `list(1, 'hi', data.frame(1:10, letters[1:10]))`
* To subset a vector or list, use the `[ ]`
  - inside the brackets: 
    + positive integer vectors for indices you want 
    + negative integer vectors for indices you want to drop
    + logical vectors for indices you want to keep/drop (TRUE/FALSE)
    + character vectors for named vectors corresponding to which named elements you want to keep
    + subsetting a list with a single square bracket always returns a list
+ To subset a list and get back just that element, use `[[ ]]`

### Exercise -- Vectors

First exercise, create a vector of ten elements. Use square brackets to subset first element, and then again for the first three elements.

```{r vecsub, exercise=TRUE}
# Create a vector of 10 numbers between 1 and 10
vec <- 1:10
# subset the first element
vec[]
# subset the first through third elements

```

```{r vecsub-solution}

# subset the first element
vec[1]
# subset the first through third elements
vec[1:3]

```


## Object Representation

+ To find the type of an object, use `class` (_higher level representation_)
+ To find how the object is stored in memory, use `typeof` (_lower level representation_)
+ To display the interal structure of an object, type `str`

### Vectors 

Use the `typeof` function to explore the data structure's low-level structure, and the `class` object to see the object's higher level structure.

```{r vectors, exercise=TRUE}
random_norms <- rnorm(100)
typeof(random_norms)
class(random_norms)

some_letters <- letters[1:10]
typeof(some_letters)
class(some_letters)

int_vector <- c(1L, 2L, 3L)
typeof(int_vector)
class(int_vector)

booleans <- int_vector == 1
typeof(booleans)
class(booleans)
```

+ can you mix types in a vector?

+ what happens when you mix the numeric with letters?

```{r coercion, exercise=TRUE}
random_norms <- rnorm(100)
some_letters <- letters[1:10]

combine_char_num <- c(random_norms, some_letters)

# check the type of combine_char_num

```

```{r coercion-solution}

# check the type of combine_char_num
typeof(combine_char_num)
```


## Lists

to combine types, make list

```{r makelist, exercise=TRUE}
random_norms <- rnorm(100)
some_letters <- letters[1:10]

list_char_num <- list(nums = random_norms, chars = some_letters)
typeof(list_char_num)
lapply(list_char_num, typeof)
```

### Exercise -- Lists

Let's see how subsetting works on lists. Let's create a list of multiple vectors. First element will be a vector of 100 random numbers from the Gaussian distribution. The next element will be a vector of letters from the built in `letters` vector. Check types and classes as you go along.

```{r list, exercise=TRUE}
my_list <- list(num = rnorm(100), chars = letters[1:10])
# subset first element using single square brackets
my_list[1]
# check it's type with typeof

# use two square brackets and select second element

# check it's type

```


```{r list-solution}

# check it's type with typeof
typeof(my_list[1])
# use two square brackets and select second element
my_list[[2]]
# check it's type
typeof(my_list[[2]])
```

## data.frames

To mix elements in a rectangular object/table, use `data.frames`:

```{r dfs, exercise=TRUE}
df_char_num <- data.frame(chars = letters[1:10], nums = rnorm(10))
lapply(df_char_num, typeof)

## data.frames must have same length in each column
```

### Working with data.frames

The elements of a data.frames are vectors.

Let's use some common vector functions like `mean` on the numeric values of a data.frame.

```{r mean, exercise=TRUE}
df_char_num <- data.frame(chars = letters[1:10], nums = rnorm(10))
# use str to see the structure of a data.frame
str(df_char_num)
# use summary for quick five-point summarizations
summary(df_char_num)
# select the first column of a data.frame
# what's the difference between
df_char_num[1]
# and
df_char_num[[1]]
# calculate the mean of the first column
```

```{r mean-solution}
mean(df_char_num[[1]])
```

